---
title: "Data Import"
output: github_document
---

```{r setup}
library(tidyverse)
library(readxl)
library(haven)
```

#### DATA IMPORT

## Read in some data

Read in the litters dataset and clean it using clean_names (e.g., turns titles in snakecase)

```{r}
litters_df = read_csv("./data/FAS_litters.csv")
litters_df = janitor::clean_names(litters_df)
```

## Take a look at data

Printing in the console, so you can see everything that is going in
```{r}
litters_df
```

Look at head/tail of the litters dataset
```{r}
tail(litters_df)
```

Gives a nice overview of the data
```{r}
skimr::skim(litters_df)
```

View the data in a separate tab using view (dataframe_name) --> but only do that in the console or else the it can mess up the knitting process

## Options to view data

Skip first row , make sure first row is not taken as names , empty values show up as NA

```{r}
litters_df = read_csv("./data/FAS_litters.csv", skip = 10, col_names = FALSE, na = c("", "NA", ".", 999))
```

Check out `?read_csv()` for more information

## Other file formats

Read in an Excel file

```{r}
mlb_df = read_excel("./data/mlb11.xlsx")
mlb_df
```

Read in a SAS file

```{r}
pulse_df = read_sas("./data/public_pulse_data.sas7bdat")
pulse_df
```

## Comparison with Base R

what about`read.csv`

```{r}
litters_base = read.csv("data/FAS_litters.csv")
litters_readr = read_csv("data/FAS_litters.csv")

litters_base
litters_readr
```


## Exporting data

Export the mlb sub-table

```{r}
write_csv(mlb_df, "./data/mlb_subtable.csv")
```


#### DATA MANIPULATION


## Load in FAS litters

```{r}
litters_df = read_csv("./data/FAS_litters.csv")
litters_df = janitor::clean_names(litters_df)
litters_df
```

## `select`

chose some columns and not others

```{r}
select(litters_df, group, gd0_weight)
```

```{r}
select(litters_df, group, gd0_weight:gd_of_birth)
```

```{r}
select(litters_df, -litter_number)
```

Renaming columns

```{r}
select(litters_df, GROUP = group, LITTer_NUMBer = litter_number)
```

```{r}
rename(litters_df, GROUP = group, LITTer_NUMBer = litter_number)
```

Select helpers

```{r}
select(litters_df, starts_with("gd"))
```

moves litter number to the beginning using two methods

```{r}
select(litters_df, litter_number, everything())
```

```{r}
relocate(litters_df, litter_number)
```

## `filter`

```{r}
filter(litters_df, gd0_weight < 22)
```

```{r}
filter(litters_df, gd0_weight >= 22)
```

```{r}
filter(litters_df, gd_of_birth == 20)
```

```{r}
filter(litters_df, !(gd_of_birth == 20))
```

same as this

```{r}
filter(litters_df, gd_of_birth != 20)
```

select multiple things

```{r}
filter(litters_df, gd0_weight >= 22, gd_of_birth == 20)
```

```{r}
filter(litters_df, group == "Mod8")
```

```{r}
filter(litters_df, group %in% c("Con7", "Mod8"))
```

## `mutate` modify existing variable or create new variable

```{r}
mutate(litters_df, wt_gain = gd18_weight - gd0_weight)
```

```{r}
mutate(
  litters_df, 
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group))
```

## `arange` order data by columns

```{r}
arrange(litters_df, pups_born_alive)
```

## `%>%`

```{r}
litters_data_raw = read_csv("./data/FAS_litters.csv")
litters_clean_name = janitor::clean_names(litters_data_raw)
litters_data_selected = select(litters_clean_name, -pups_survive)
litters_mutated = mutate(litters_data_selected, wt_gain = gd18_weight - gd0_weight)
litters_without_missing = drop_na(litters_mutated, gd0_weight)
```

USE THE PIPE OPERATOR INSTEAD

```{r}
litters_df = 
  read_csv("./data/FAS_litters.csv") %>% 
  janitor::clean_names() %>% 
  select(-pups_survive) %>% 
  mutate(wt_gain = gd18_weight - gd0_weight) %>% 
  drop_na(gd0_weight)
```



#### TIDY DATA


## pivot_longer

```{r}
pulse_df =
  haven::read_sas("./data/public_pulse_data.sas7bdat") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = 'visit',
    values_to = 'bdi',
    names_prefix = 'bdi_score_'
  ) %>% 
mutate(
  visit = replace(visit, visit == 'bl', '00m'),
  visit = factor(visit)
) %>% 
arrange(id, visit)
```

## learning assessment 1

```{r}
litters_wide = 
  read_csv("./data/FAS_litters.csv") %>%
  janitor::clean_names() %>%
  select(litter_number, ends_with("weight")) %>% 
  pivot_longer(
    gd0_weight:gd18_weight,
    names_to = "gd", 
    values_to = "weight") %>% 
mutate(gd = recode(gd, "gd0_weight" = 0, "gd18_weight" = 18))
```

## pivot_wider

```{r}
analysis_result = tibble(
  group = c("treatment", "treatment", "placebo", "placebo"),
  time = c("pre", "post", "pre", "post"),
  mean = c(4, 8, 3.5, 4)
)
```

```{r}
analysis_result_wide = pivot_wider(
  analysis_result, 
  names_from = "time", 
  values_from = "mean")
```


## binding rows

Open Excel to make this easier (i.e., knowing which columns to pick). First step is to create 3 separate tables, one for each movie

```{r}
fellowship_ring = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "B3:D6") %>%
  mutate(movie = "fellowship_ring")

two_towers = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "F3:H6") %>%
  mutate(movie = "two_towers")

return_king = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "J3:L6") %>%
  mutate(movie = "return_king")
```

Now we combine the three tables.

```{r}
lotr_tidy = 
  bind_rows(fellowship_ring, two_towers, return_king) %>%
  janitor::clean_names() %>%
  pivot_longer(
    female:male,
    names_to = "gender", 
    values_to = "words") %>%
  mutate(race = str_to_lower(race)) %>% 
  select(movie, everything()) 
```

## Joining datasets

```{r}
pups = 
  read_csv("./data/FAS_pups.csv") %>%
  janitor::clean_names() %>%
  mutate(
    sex = recode(sex, `1` = "male", `2` = "female"),
    sex = factor(sex)) 
```

```{r}
litter = 
  read_csv("./data/FAS_litters.csv") %>%
  janitor::clean_names() %>%
  separate(group, into = c("dose", "day_of_tx"), sep = 3) %>%
  relocate(litter_number) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    dose = str_to_lower(dose))
```

Now we do a left-join because we want to analyze the data of the babies (not the mothers)

```{r}
fas_data = 
  left_join(pups, litter, by = "litter_number")
```

## learning assessment 2

```{r}
surv_os = read_csv("data/surv_os.csv") %>% 
  janitor::clean_names() %>% 
  rename(id = what_is_your_uni, os = what_operating_system_do_you_use)

surv_pr_git = read_csv("data/surv_program_git.csv") %>% 
  janitor::clean_names() %>% 
  rename(
    id = what_is_your_uni, 
    prog = what_is_your_degree_program,
    git_exp = which_most_accurately_describes_your_experience_with_git)

surv_data = 
  full_join(surv_os, surv_pr_git, by = "id")
```





