---
title: "Homework 1 - P8105"
author: "Tim Hauser (th2899)"
date: 2022-09-23
output: github_document
---

```{r setup, include = FALSE}
library(tidyverse)
```

The purpose of this file is to document the steps to fulfill Homework 1. 

## Problem 1

Load the Penguins dataset

```{r load_data}
data("penguins", package = "palmerpenguins")
```

The `r nrow(penguins)` (rows) by `r ncol(penguins)` (columns) penguins dataset is a list of penguins with the following eight variables collected for each (if available): `r ls(penguins)`.

The mean flipper length across all penguins in the data set is `r round(mean(penguins$flipper_length_mm, na.rm=TRUE), 3)` mm. As a side note, in the mean calculation I ignored the NA value in row 4, because R by default returns a mean of NA if the data contains at least one NA.

Below is displayed a scatterplot with `flipper_length_mm` on the Y axis and `bill_length_mm` on the X axis, colored by `species`:

```{r yx_scatter, warning=FALSE}
ggplot(penguins, aes(x = bill_length_mm, y = flipper_length_mm, color = species)) + geom_point()
```

```{r save_plot, warning=FALSE}
ggsave("scatter_plot.pdf", height = 4, width = 6) 
```



## Problem 2

The following code creates the data frame as instructed (I set a seed for reproducibility `set.seed(1234)`):

```{r df_create}
set.seed(1234)

p2_df = tibble(
  var_numeric = rnorm(10),
  var_logic = c(ifelse(var_numeric>0, "TRUE", "FALSE")),
  var_character = c("This", "really", "is", "an", "exemplary", "character", "vector", "of", "length", "ten"),
  var_factor = factor(c("male", "female", "male", "male", "undefined", "male", "female", "male", "male", "undefined"))
)
```

Now we will take the mean of each variable created in above dataframe to see what works and what does not:

* Numeric: `r round(mean(pull(p2_df, 1)), 3)` --> the mean function works as expected, since the frame contains numerical values (btw, another was of solving this is code: `r round(mean(p2_df$var_numeric), 3)`)
* Logic: `r round(mean(pull(p2_df, 2)), 3)` --> the mean function does not work on the logic variable, R does not seem to know how to handle it
* Character: `r round(mean(pull(p2_df, 3)), 3)` --> the mean function does not work on the logic variable, R does not seem to know how to handle it
* Factor: `r round(mean(pull(p2_df, 3)), 3)` --> the mean function does not work on the logic variable, R does not seem to know how to handle it

Next, we try let `as.numeric` work on our dataframe to see if that solves the issues occurring above for logic, character, and factor variables. We do this by creating a new dataframe based on the old one but transforming the logic, character and factor variable columns into numeric values:

```{r, include=TRUE, warning=FALSE}
p2_df_2 = transform(p2_df, var_logic = as.numeric(var_logic), var_character = as.numeric(var_character), var_factor = as.numeric(var_factor))
```

* Logic: `r round(mean(pull(p2_df_2, 2)), 3)` --> the  `as.numeric` function converted the logic values into NA (due to coercion), hence still no mean could be calculated 
* Character: `r round(mean(pull(p2_df_2, 3)), 3)` --> the  `as.numeric` function converted the character values into NA (due to coercion), hence still no mean could be calculated 
* Factor: `r round(mean(pull(p2_df_2, 4)), 3)` --> the  `as.numeric` function converted successfully converted the factor values into numeric ones, it converted the alphabetically first appearing unique value into a 1 (female), the second into a 2 (male), and the third into a 3(undefined), since the column has six males, two females, and two undefined values, the mean must be 2. 